Overview

### 2023-11-21 analysis/001_overview.Rmd
Analyse five-years period

```{r overview_data_prep}
ta_oa_growth <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", cr_year != "2023") |>
  group_by(cr_year) |>
  summarise(ta_oa = n_distinct(doi))
hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", cr_year != "2023") |>
  group_by(cr_year) |>
  summarise(oa_all = n_distinct(doi))
ta_df <- inner_join(ta_oa_growth, hybrid_oa_growth) |>
  mutate(ta_oa_prop = ta_oa / oa_all) |>
  mutate(no_ta = oa_all - ta_oa, 
         no_ta_prop = 1 - ta_oa_prop) |>
  mutate(cr_year = as.factor(cr_year))
# year all
year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", cr_year != "2023") |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year) |> 
  summarise(n = sum(jn_all))

## Hybrid Journals
 n_hybrid_jns_all <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", cr_year != "2023") |>
  distinct(issn_l) |>
   nrow()

## Hybrid Journals with at least one OA article
n_hybrid_jns <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", cr_year != "2023") |>
  group_by(issn_l) |>
  summarise(oa_all = n_distinct(doi)) |> 
  nrow()

## Hybrid Journals with at least one OA article
n_ta_jns <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", cr_year != "2023") |>
  distinct(issn_l) |>
  nrow()

## Hybrid Journals with at least one TA-OA article
n_tas_with_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", cr_year != "2023") |>
  distinct(ta_journal_portfolio) |>
  nrow()

## Open access volume
n_oa <-  hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", cr_year != "2023") |>
  distinct(doi) |>
  nrow()

## Total
n_total <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", cr_year != "2023", !is.na(cc)) |>
  distinct(issn_l, cr_year, jn_all) |>
  pull(jn_all) |>
  sum()

## TA total
n_tas_with_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", cr_year != "2023") |>
  distinct(ta_journal_portfolio) |>
  nrow()

## TA article total
n_ta_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", cr_year != "2023") |>
  distinct(doi) |>
  nrow()
```

Between 2017 and 2022, a total of `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements provided at least one open access article under a Creative Commons license.
During this period, `r n_oa` out of `r n_total` articles in these journals became open access, representing a six-year hybrid open access uptake rate of `r n_oa / n_total * 100`%.
First authors affiliated with institutions that had a transformative agreement in place at the time of publication, which indicates that they could make use of such agreements, contributed `r n_ta_oa` open access articles to the total.

```{r area_oa_over_the_years}
oa_ta_plot <- ta_df |>
  select(cr_year, ta_oa, no_ta) |>
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  filter(cr_year != "2023") |>
  inner_join(year_all) |>
  # plot
  ggplot(aes(cr_year, value / n, fill = type, group = type)) +
  geom_area() +
  scale_fill_manual("Hybrid OA", values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement"),  guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::percent_format(), limits = c(0, .15), breaks = c(0, .05, .1, .15)) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(legend.position="top", legend.justification = "left", axis.text.x = element_blank())
```

```{r ta_share_plot}
ta_share_plot <- ta_df |>
  select(cr_year, ta_oa_prop, no_ta_prop) |>
  pivot_longer(cols = c(ta_oa_prop, no_ta_prop), names_to = "type", values_to = "value") |>
  filter(cr_year != "2023", type == "ta_oa_prop") |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value, group = 1)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = c(0, 0.25, 0.5),
                     labels = scales::percent) +
  theme_minimal() +
  labs(y = "Percentage\nof Hybrid OA\nvia Agreement", x = "Publication Year")
```

@@

oa_jn_box <- left_join(jn_hybrid_oa_growth, jn_year_all) |>
  mutate(prop = oa_all / n) |> 
  filter(cr_year != "2023") |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), prop)) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  coord_cartesian(ylim = c(0,0.65)) +
  theme_minimal() +
  labs(y = "OA Uptake per Hybrid Journal", x = "Publication Year")
```

@@

summary_all <- inner_join(ta_df, jn_year_all, by = c("cr_year", "issn_l")) |>
  mutate(prop_ta_oa = ta_oa / n,
         prop_no_ta = no_ta / n) |>
  ungroup() |>
  distinct(cr_year, issn_l, prop_ta_oa, prop_no_ta) |>
  pivot_longer(cols = c(prop_ta_oa, prop_no_ta))
# plot
box_ta_prop <- summary_all |>
  filter(cr_year != "2023") |>
  # With at least one OA article
  filter(value > 0) |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value, fill = name)) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  coord_cartesian(ylim = c(0,0.65)) +
  scale_fill_manual("Hybrid OA", values = c(prop_no_ta = "#b3b3b3a0", prop_ta_oa = "#56B4E9"), labels = c(prop_no_ta = "Other", prop_ta_oa = "Agreement"), guide = FALSE) +
  theme_minimal() +
  labs(y = NULL, x = "Publication Year")

@@

## plot
publisher_plot_df <- publisher_df |> 
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  mutate(
    publisher = case_when(
      esac_publisher == "Elsevier" ~ "Elsevier",
      esac_publisher == "Springer Nature" ~ "Springer Nature",
      esac_publisher == "Wiley" ~ "Wiley",
      is.character(esac_publisher) ~ "Other"
    )
  ) |>
  mutate(publisher =
           forcats::fct_relevel(
             publisher,
             c("Elsevier", "Springer Nature", "Wiley", "Other") 
           )) |>
  filter(cr_year != "2023", oa > 0)

>>>>

```{r overview_data_prep}
ta_oa_growth <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(ta_oa = n_distinct(doi))
hybrid_oa_growth <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(cr_year) |>
  summarise(oa_all = n_distinct(doi))
ta_df <- inner_join(ta_oa_growth, hybrid_oa_growth) |>
  mutate(ta_oa_prop = ta_oa / oa_all) |>
  mutate(no_ta = oa_all - ta_oa, 
         no_ta_prop = 1 - ta_oa_prop) |>
  mutate(cr_year = as.factor(cr_year))
# year all
year_all <-
  hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(issn_l, cr_year, jn_all) |> 
  group_by(cr_year) |> 
  summarise(n = sum(jn_all))

## Hybrid Journals
 n_hybrid_jns_all <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(issn_l) |>
   nrow()

## Hybrid Journals with at least one OA article
n_hybrid_jns <- hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  group_by(issn_l) |>
  summarise(oa_all = n_distinct(doi)) |> 
  nrow()

## Hybrid Journals with at least one OA article
n_ta_jns <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(issn_l) |>
  nrow()

## Hybrid Journals with at least one TA-OA article
n_tas_with_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(ta_journal_portfolio) |>
  nrow()

## Open access volume
n_oa <-  hoaddata::cc_articles |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(doi) |>
  nrow()

## Total
n_total <- hoaddata::jn_ind |>
  filter(issn_l != "0027-8424", !cr_year %in%  c("2017", "2023"), !is.na(cc)) |>
  distinct(issn_l, cr_year, jn_all) |>
  pull(jn_all) |>
  sum()

## TA total
n_tas_with_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(ta_journal_portfolio) |>
  nrow()

## TA article total
n_ta_oa <- ta_oa_inst |>
  filter(ta_active == TRUE, !is.na(cc), issn_l != "0027-8424", !cr_year %in%  c("2017", "2023")) |>
  distinct(doi) |>
  nrow()
```

Between 2017 and 2022, a total of `r n_hybrid_jns` out of `r n_hybrid_jns_all` hybrid journals in transformative agreements provided at least one open access article under a Creative Commons license.
During this period, `r n_oa` out of `r n_total` articles in these journals became open access, representing a five-year hybrid open access uptake rate of `r n_oa / n_total * 100`%.
First authors affiliated with institutions that had a transformative agreement in place at the time of publication, which indicates that they could make use of such agreements, contributed `r n_ta_oa` open access articles to the total.

```{r area_oa_over_the_years}
oa_ta_plot <- ta_df |>
  select(cr_year, ta_oa, no_ta) |>
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  inner_join(year_all) |>
  # plot
  ggplot(aes(cr_year, value / n, fill = type, group = type)) +
  geom_area() +
  scale_fill_manual("Hybrid OA", values = c(no_ta = "#b3b3b3a0", ta_oa = "#56B4E9"), labels = c(no_ta = "Other", ta_oa = "Agreement"),  guide = guide_legend(reverse = TRUE)) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), labels = scales::percent_format(), limits = c(0, .15), breaks = c(0, .05, .1, .15)) +
  theme_minimal() +
  labs(y = "Hybrid OA Uptake", x = NULL) +
  theme(legend.position="top", legend.justification = "left", axis.text.x = element_blank())
```

```{r ta_share_plot}
ta_share_plot <- ta_df |>
  select(cr_year, ta_oa_prop, no_ta_prop) |>
  pivot_longer(cols = c(ta_oa_prop, no_ta_prop), names_to = "type", values_to = "value") |>
  filter(!cr_year %in%  c("2017", "2023"), type == "ta_oa_prop") |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value, group = 1)) +
  geom_bar(stat = "identity", fill = "#56B4E9") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)), breaks = c(0, 0.25, 0.5),
                     labels = scales::percent) +
  theme_minimal() +
  labs(y = "Percentage\nof Hybrid OA\nvia Agreement", x = "Publication Year")
```

@@

oa_jn_box <- left_join(jn_hybrid_oa_growth, jn_year_all) |>
  mutate(prop = oa_all / n) |> 
  filter(!cr_year %in%  c("2017", "2023")) |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), prop)) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  coord_cartesian(ylim = c(0,0.65)) +
  theme_minimal() +
  labs(y = "OA Uptake per Hybrid Journal", x = "Publication Year")
```

@@

summary_all <- inner_join(ta_df, jn_year_all, by = c("cr_year", "issn_l")) |>
  mutate(prop_ta_oa = ta_oa / n,
         prop_no_ta = no_ta / n) |>
  ungroup() |>
  distinct(cr_year, issn_l, prop_ta_oa, prop_no_ta) |>
  pivot_longer(cols = c(prop_ta_oa, prop_no_ta))
# plot
box_ta_prop <- summary_all |>
  filter(!cr_year %in%  c("2017", "2023")) |>
  # With at least one OA article
  filter(value > 0) |>
  ggplot(aes(gsub("^20", "'", as.character(cr_year)), value, fill = name)) +
  geom_boxplot(outlier.shape = NA) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.05)),
                     labels = scales::percent) +
  coord_cartesian(ylim = c(0,0.65)) +
  scale_fill_manual("Hybrid OA", values = c(prop_no_ta = "#b3b3b3a0", prop_ta_oa = "#56B4E9"), labels = c(prop_no_ta = "Other", prop_ta_oa = "Agreement"), guide = FALSE) +
  theme_minimal() +
  labs(y = NULL, x = "Publication Year")

@@

## plot
publisher_plot_df <- publisher_df |> 
  pivot_longer(cols = c(ta_oa, no_ta), names_to = "type", values_to = "value") |>
  mutate(
    publisher = case_when(
      esac_publisher == "Elsevier" ~ "Elsevier",
      esac_publisher == "Springer Nature" ~ "Springer Nature",
      esac_publisher == "Wiley" ~ "Wiley",
      is.character(esac_publisher) ~ "Other"
    )
  ) |>
  mutate(publisher =
           forcats::fct_relevel(
             publisher,
             c("Elsevier", "Springer Nature", "Wiley", "Other") 
           )) |>
  filter(!cr_year %in%  c("2017", "2023"), oa > 0)
